---
title: "Class 8: PCA exercise"
output: html_notebook
---

# Principal component analysis with food in UK

Read in our data file, view what we are working with 

```{r}

# set the index of the dataframe as the first row
w <- read.csv("UK_foods.csv", row.names = 1)

w

```

How many rows and columns 
```{r}

dim(w)

```

Checking the top of the data
```{r}

head(w)

```

Start visulaizing data

```{r}

barplot(as.matrix(w), beside=T, col=rainbow(nrow(w)))

```

Changing beside to false gives you a much different plot

```{r}

barplot(as.matrix(w), beside=F, col=rainbow(nrow(w)))

```

Generating pairwise plots to compare 
```{r}

pairs(w, col=rainbow(10), pch=16)
 

```

Principle Component Analysis makes comparing these multivariable datasets easier

R makes this easy by using the prcomp() function

```{r}

pca <- prcomp( t(w))
summary(pca)

```

This summary shows us that PC1 accounts for 67% of the sample variance.  

We can plot principle components against each other:

```{r}

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(w))

```

We can jazz up this plot by adding colors to our labels

```{r}

my_cols <- c("orange", "red","blue", "green")

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500), col = my_cols)

text(pca$x[,1], pca$x[,2], colnames(w), col= my_cols)

```

Below we can use the square of pca$sdev , which stands for “standard deviation”, to calculate how much variation in the original data each PC accounts for.

```{r}

v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v

```

We can plot principle components by the amount of variance they contribute
```{r}

barplot(v, xlab="Principal Component", ylab="Percent Variation")

```

Can plot across one principle component (the one with the most variance) tell which of the variables are contributing the most

```{r}

## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(8, 4, 4, 3))
barplot( pca$rotation[,1], las=2, cex.names = .75)

```

Now do the same with PC2
```{r}

## Now plot PC2
par(mar=c(8, 4, 4, 3))
barplot( pca$rotation[,2], las=2, cex.names = .75)

```

We can also try to plot a biplot which are popular ways to represent PCA data

```{r}

biplot(pca)

```

# Section 2: PCA of RNA-seq data

Read in RNA-seq data 

```{r}
mydata <- read.csv("expression.csv", row.names = 1)

dim(mydata)

pca <- prcomp(t(mydata), scale=TRUE) 
attributes(pca)

plot(pca$x[,1], pca$x[,2]) 
pca.var <- pca$sdev^2 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1) 
pca.var.per

xlab <- paste("PC1 (", pca.var.per[1], %)", sep="")
ylab <- paste("PC2 (", pca.var.per[2], %)", sep="")

xlab 
ylab
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")


## A vector of colors for wt and ko samples
colvec <- as.factor( substr( colnames(mydata), 1, 2) )
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
```


