---
title: "Class 14: RNA-seq"
output: github_document
---

# Section 1: Introduction

In this exercise we are going to analyze data from a published RNA-seq dataset. In this dataset they looked at RNA-seq gene expression data in 4 airway smooth muscle cells with dexamethasone glucocorticoid and without. 
 
First install relevant packages 

```{r}

#install.packages("BiocManager")
#BiocManager::install()

#BiocManager::install("DESeq2")

```

Then load the relevant packages 

```{r}

library(BiocManager)
library(DESeq2)

```

 
# Section 2: Import data

First read.csv() function to read these count data and metadata files.

```{r }


counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)

metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)

```

View the data 

```{r }

head(counts)

nrow(counts)


```

What does the metadata have in it?
```{r}

head(metadata)

```

# Section 3: Go through steps of differential gene expression (normally will do DESEq2)
```{r}

# pluck out the 4 cell line names that are controls
controls = (metadata$dex == "control")
control_lines = metadata$id[controls]
control_lines

#control.mean = rowSums(counts[, control_lines$id])/4

#head(control_counts)
```

Find the samples that are controls, calculate counts per gene for these samples. 

```{r}

control <- metadata[metadata[,"dex"]=="control",]

control.mean <- rowSums( counts[ ,control$id] )/length(control$id)

names(control.mean) <- counts$ensgene

```

Do the same thing with the treated samples 
```{r}

treated <- metadata[metadata[,"dex"]=="treated",]

treated.mean <- rowSums( counts[ ,treated$id] )/length(treated$id)

names(treated.mean) <- counts$ensgene


```


Combine these together into a new data.frame called 'meancounts'

```{r}

meancounts <- data.frame(control.mean, treated.mean)
meancounts

```

Creating a scatter plot comparing the mean of treated sample versus the mean of control samples. 

```{r}

plot(control.mean, treated.mean)

```

Plot the log of each instead
```{r}

plot(log(control.mean), log(treated.mean))

```


Here we calculate log2foldchange, add it to our meancounts data.frame. Log2foldchange is frequently used to compare expression between treated and control.

```{r}

meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])

head(meancounts)

```

When we look at our dataset we have values that do not make sense. We see NaN from dividing by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Filter our data to remove these genes. 

```{r}

# find the zeros
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

# remove the zeros
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]

# check to see if it worked
head(mycounts)

```

A log fold change of absolute value two is considered to be a significant enough change. We will pull these genes out of our data set. 

```{r}

up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)

```

how many upregulated?
```{r}

sum(up.ind)

```

how many downregulated?
```{r}

sum(down.ind)

```

# Section 4: Adding annotation data 

Our data only identifies genes by Ensembl IDs, we want to add additional annotation data to make our results more understandable 

Here we will read in an excel file of annotation information:

```{r}

anno <- read.csv("data/annotables_grch38.csv")

head(anno)

```

Ideally we want this annotation data mapped (or merged) with our mycounts data. 
```{r}

merge(x = mycounts, y= anno, by.x =  "row.names", by.y = "ensgene")

```

You can also use Bioconductor annotation packages. Here we load the AnnotationDbi package and the annotation package org.Hs.eg.db.

```{r}

library("AnnotationDbi")
library("org.Hs.eg.db")

```

Look at the types of annotations that are available
```{r}

columns(org.Hs.eg.db)

```

We can use the mapIds() function to add individual columns to our results table. 
keys is  where our genenames are located 
keytype is the type of annotation our genenames are in
column tells us the format from annotation file we are adding

```{r}

mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), 
      
                     keytype="ENSEMBL",        
                     
                     column="SYMBOL",          
                    
                     multiVals="first")

```

```{r}

# check if our column added 

head(mycounts)

```

Repeat the mapIds and this time add ENTREZID
```{r}

mycounts$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), 
                     
                     keytype="ENSEMBL",        
                     
                     column="ENTREZID",   
                     
                     multiVals="first")

```

Do the same thing but add UNIPROT annotations
```{r}

mycounts$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), 
                     
                     keytype="ENSEMBL",     
                     
                     column="UNIPROT",     
                     
                     multiVals="first")

```

```{r}

# did it work? yes! 
head(mycounts)

```

Check our upregulated data
```{r}

#View(mycounts[up.ind,])

```






