---
title: "Class 4: R Basics"
output: github_document
---

In-Class exercise as a part of BGGN213 W19 at UCSD
Instructor: Barry Grant

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 2: Simple Calculations

Starting with attempting basic arthmetic 

Can do addition:
```{r }

5 + 3

```


Subtraction:
```{r }


19  -4


```

Multiplication
```{r}

4 * 7 


```


Division 
```{r}

15 / 3



```


Exercise 3: Saving your answers using object assignmnet


Assign the result of a calculation to an object.
View object.
```{r}

x <- 2 * 7 * 9
x

# NOTE use <- and not = which will cause issues later

```


Exercise 4: Calling Functions 


R has a lot of built in functions that can make calculations quick and easy. Here are a few examples. 

```{r}

seq(1,10)

```

If you don't know how a function works then you can use help(function) to get information about it

```{r}

help(seq)

```

Using additional arguments to seq

```{r}

seq(1,10, by =2)

```

Exercise 5: How to get help in R

```{r}

help(log)
?log

```

If you don't know which function you need you can do the following

```{r}

help.search("cross tabulate")
??"cross tabulate"

```


Each function has an example file

```{r}

example(log)

```


Excercise 6: Vectors and Indexing

Vectors are a means for storing data in R

```{r}

length(3.1)

```


To combine values into vectors use c()

```{r}

x <- c(121, 4.56, 8.235)
x

```


```{r}

y <- c(898, 1.543, 705)
y

```


Vectorization means that R loops over the values in vectors. Allows us to use arithmetic functions to vectors of the same length.

```{r}

x + y


```

```{r}

x - y 

```

```{r}

x / y 

```

```{r}

sqrt(x)

```

```{r}

round(sqrt(x), 3)

```

```{r}

log(x)/2 + 1

```

Vector Indexing: We can use indexing to access specific values within a vector. 

```{r}

x <- c(56, 95.3, 0.4)
x[2]

```

Index 1 responds to the first value in a vector 


```{r}

x[1]

```

Note on reproducibility

Use sessionInfo() to record the version of R and the packages used to make sure everything is reproducible

```{r}

sessionInfo()

```

