---
title: 'Class 7: R functions and packages'
output: github_document
---

#This is work from an in-class exercise in function writing. 

First we will use the same function from the other day.

# Function 1: Rescale

```{r}
source("http://tinyurl.com/rescale-R")

# insert a stop statment to fail early and loudly 

rescale <- function(x, na.rm=TRUE, plot=FALSE, ...) {
  if( !is.numeric(x) ) {
 stop("Input x should be numeric", call.=FALSE)
 }
  rng <-range(x, na.rm=na.rm)
  
  answer <- (x - rng[1]) / (rng[2] - rng[1])
  if(plot) {
    plot(answer, ...)
  }
  return(answer)
}

rescale( c(1:10) )

```

# Function 2: both_na

We want to write a function, called both_na(), that counts how many positions in two input vectors, x and y, both have a missing value

```{r}

# start with a smaller problem, tell if there is any NA in both functions

# give your function small input to test if it works 

x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)


if (sum(is.na(x)) >= 1 & sum(is.na(y)) >= 1) {
  print("I work!")
}


```

```{r}

# now put this into a function 

both_na <- function(x,y) {
  if (sum(is.na(x)) >= 1 & sum(is.na(y)) >= 1) {
  print("I work!")
}
}


```

Now that our small function works we can make it more complicated! 

```{r}

both_na3 <- function(x, y) {
 if(length(x) != length(y)) {
   # add in sanity checks to prevent bugs
 stop("Input x and y should be vectors of the same length")
 }

 na.in.both <- ( is.na(x) & is.na(y) )
 na.number <- sum(na.in.both)
 na.which <- which(na.in.both)
 message(
  
 )
# return relevant data we want from our function 
 return( list(number=na.number, which=na.which) )
}


both_na3(x,y)
```

# Function 3: Calculate Students' Grades 

Calculate a function to output the average grade of a student. For this function I turned any NAs to 0s. Also drop the lowest grade for each student when calculating the average. 

```{r}

#here is our data input we are working with 

student1 <- c(100, 100, 100, 100, 100,100, 100, 90)
student2 <- c(100, NA, 90, 90,90, 90, 97, 80)

# here is our function 
calculate_grade <- function(student){
  # take NAs and convert them to 0s
  if(sum(is.na(student)) >= 1) {
    student <- replace(student, is.na(student), 0)
  }
  # find the lowest grade for the student
  this_min = min(student)
  # drop this lowest grade 
  average = (sum(student) - this_min) / (length(student) - 1)
  # print out the results 
  print("The average for the student is:")
  print(average)
}

# run my function
calculate_grade(student2)

```

Testing our my new function
```{r}
# read in a file of student scores 
df <- read.csv('student_homework.csv', row.names = 1)
df

# run my function on my file 
calculate_grade(df[1,])

# try using apply to run my function on the file
answer <- apply(df, 1 ,calculate_grade)

```

```{r}

# sort the students by their grades

sort(answer, decreasing = TRUE)

```

# Function 4: Find common genes in two datasets 

This is another example problem from the class 7 lecture. 

We will start small with a example dataset.

```{r}

# create two data frames with genes and their expression 

df1 <- data.frame(IDs=c("gene1", "gene2", "gene3"),
 exp=c(2,1,1),
 stringsAsFactors=FALSE)

df2 <- data.frame(IDs=c("gene2", "gene4", "gene3", "gene5"),
 exp=c(-2, NA, 1, 2),
 stringsAsFactors=FALSE)

```


Simplify these into single vectors 
```{r}

x <- df1$IDs
y <- df2$IDs

```


We can use the built in function to R of intersect
```{r}

intersect(x, y)

# this isn't quite what we want we want their locations or indicies
```


```{r}

# instead let's try 

x %in% y

```

Find the smallest working snippet that you can then put into a function
```{r}

cbind( x[ x %in% y ], y[ y %in% x ] )

```

Now that it works put it into a function! Woot!
```{r}

gene_intersect <- function(x, y) {
  cbind( x[ x %in% y ], y[ y %in% x ] )
}

```

Now run our function:
```{r}

gene_intersect(x,y)

```

# Exercise to practice using packages in R. We practiced used ggplot2

First install like this:
install.packages("BiocManager")
install.packages("ggplot2")
library("ggplot2")


```{r}

# here we are reading in a library with data

# example of how plotting in base R

library(tidyverse)

housing <- read_csv("dataSets/landdata-states.csv")

plot(Home.Value ~ Date,
     col = factor(State),
     data = filter(housing, State %in% c("MA", "TX")))

legend("topleft",
       legend = c("MA", "TX"),
       col = c("black", "red"),
       pch = 1)


```

```{r}

# example of how using ggplot it becomes a much more beautiful plot! 

ggplot(filter(housing, State %in% c("MA", "TX")),
       aes(x=Date,
           y=Home.Value,
           color=State))+
  geom_point()


```

```{r}

sessionInfo()

```



